---
title: "data_cleaning"
output: html_document
date: "2025-01-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Step 2: Data Cleaning 

# 2A: Cleaning the Mouse Files 
## Validation 
Here, we ensured each element in the columns matched the description outlined in the SOP file. 

### Validate string length ( 15 characters) in analysis id

```{r}
# Create a loop
validate_analysis_id <- function (x, dataType, min_val, max_val) { # Validate length of elements in the vector 'x'
  if (tolower(dataType) == "string") { # Check if the `dataType` (converted to lowercase) is a string
    invalid_analysis_id <- which(sapply(x, nchar) < min_val | sapply (x, nchar) > max_val) # If the string length is less than the minimum value, or more than the maximum value, the result was stored in invalid_analysis_id
    return(invalid_analysis_id) # Returns the invalid elements
  } 
}
  
# Apply the function to analysis_id 
invalid_analysis_id <- validate_analysis_id(final_combined_data$analysis_id, "string", 15, 15) # In the final combined data, looking at the analysis_id column, make sure the elements have a string length with a minimum and maximum value of 15 

# Print results, if there are any invalid elements
print(invalid_analysis_id)

# The returned results show 0 invalid analysis ids. This means that all strings are valid, and have the lengths defined in the SOP file 
```

### Validate string length for gene accession id

```{r}
# Create a loop
validate_gene_accession_id <- function (x, dataType, min_val, max_val) { # Validate length of elements in the vector 'x'
  if (tolower(dataType) == "string") { # Check if the `dataType` (converted to lowercase) is a string
    invalid_gene_accession_id <- which(sapply(x, nchar) < min_val | sapply (x, nchar) > max_val)  # If the string length is less than the minimum value, or more than the maximum value, the result was stored in invalid_gene_accession_id
    return(invalid_gene_accession_id) # Returns the invalid elements
  } 
}
  
# Apply the function to gene_accession_id
invalid_gene_accession_id <- validate_gene_accession_id(final_combined_data$gene_accession_id, "string", 9, 11) # In the final combined data, looking at the gene_accession_id column, make sure the elements have a string length with a minimum value of 9, and maximum value of 11

# Print results 
print(invalid_gene_accession_id)

# The returned results show 0 invalid gene accession ids. This means that all strings are valid, and have the lengths defined in the SOP file

```

### Validate string length for gene_symbol

```{r}
# Create a loop
validate_gene_symbol <- function (x, dataType, min_val, max_val) {  # Validate length of elements in the vector 'x'
  if (tolower(dataType) == "string") { # Check if the `dataType` (converted to lowercase) is a string
    invalid_gene_symbol <- which(sapply(x, nchar) < min_val | sapply (x, nchar) > max_val)  # If the string length is less than the minimum value, or more than the maximum value, the result was stored in invalid_gene_symbol
    return(invalid_gene_symbol) # Returns the invalid elements
  } 
}
  
# Apply the function to gene_symbol 
invalid_gene_symbol <- validate_gene_symbol(final_combined_data$gene_symbol, "string", 1, 13) # In the final combined data, looking at the gene_symbol column, make sure the elements have a string length with a minimum value of 1, and maximum value of 13

#Print results 
print(invalid_gene_symbol)

# The returned results show 0 invalid gene_symbols. This means that all strings are valid, and have the lengths defined in the SOP file

# The elements in the gene_symbol column have inconsistent case (some are lowercase, whereas others are uppercase) 
# Normalise gene symbols to lowercase because SQL cannot handle case sensitive information 
final_combined_data$gene_symbol <- tolower(final_combined_data$gene_symbol)
normalised_gene_symbol <- final_combined_data

# Check the first few rows to confirm the transformation
head(normalised_gene_symbol)

# Save the output file
write.csv(normalised_gene_symbol, "~/Desktop/DCDM_GRP7/outputs/normalised.csv", row.names = FALSE)
```

### Validate string length for parameter_id

```{r}
# Create a loop
validate_parameter_id <- function (x, dataType, min_val, max_val) { # Validate length of elements in the vector 'x'
  if (tolower(dataType) == "string") { # Check if the `dataType` (converted to lowercase) is a string
    invalid_parameter_id <- which(sapply(x, nchar) < min_val | sapply (x, nchar) > max_val)  # If the string length is less than the minimum value, or more than the maximum value, the result was stored in invalid_parameter_id
    return(invalid_parameter_id) # Returns the invalid elements
  } 
}
  
# Apply the function to the parameter_id 
invalid_parameter_id <- validate_parameter_id (normalised_gene_symbol$parameter_id, "string", 15, 18)  # In the final combined data, looking at the parameter_id column, make sure the elements have a string length with a minimum value of 15, and maximum value of 18

#Print results 
print(invalid_parameter_id)

# The returned results show 0 invalid parameter ids. This means that all strings are valid, and have the lengths defined in the SOP file
```

### Validate string length for parameter_name

```{r}
# Create a loop
validate_parameter_name <- function (x, dataType, min_val, max_val) { # Validate length of elements in the vector 'x'
  if (tolower(dataType) == "string") { # Check if the `dataType` (converted to lowercase) is a string
    invalid_parameter_name <- which(sapply(x, nchar) < min_val | sapply (x, nchar) > max_val)  # If the string length is less than the minimum value, or more than the maximum value, the result was stored in invalid_parameter_name
    return(invalid_parameter_name) # Returns the invalid elements
  } 
}
  
# Apply the function to the parameter_name
invalid_parameter_name <- validate_parameter_name(normalised_gene_symbol$parameter_name, "string", 2, 74) # In the final combined data, looking at the parameter_name column, make sure the elements have a string length with a minimum value of 2, and maximum value of 74

#Print results 
print(invalid_parameter_name)

# The returned results show 0 invalid parameter names. This means that all strings are valid, and have the lengths defined in the SOP file
```

### Validate the mouse life stage to the 7 categories outlined in the SOP file 

```{r}
# Create a loop 
validate_mouse_life_stage <- function (x, field_name) { # Validate column names in the vector 'x'
  if (field_name == "mouse_life_stage") {
    valid_mouse_life_stage <- c("E12.5", "E15.5", "E18.5", "E9.5", "Early adult", "Late adult", "Middle aged adult") # Define a vector containing all valid mouse life stages
    invalid_mouse_life_stage <- which(!(x %in% valid_mouse_life_stage)) # If the mouse life stages do not match the seven categories, the result was stored in invalid_mouse_life_stage
    return(invalid_mouse_life_stage) # Returns the invalid elements
  }
}

# Apply the function to the mouse_life_stage
invalid_mouse_life_stage <- validate_mouse_life_stage(normalised_gene_symbol$mouse_life_stage, "mouse_life_stage") # In the final combined data, looking at the parameter_name column, make sure the elements match a life stage out of the seven 

#Print the invalid rows 
print(invalid_mouse_life_stage)

# Print the total number of invalid mouse strains
cat("Total number of invalid mouse life stages:", length(invalid_mouse_life_stage))

# The results show there are 0 entries with invalid life stages
```

### Validate the range of the float p-value

```{r}
# Create a loop 
validate_p_value <- function (x, min_val, max_val) { # Validate column values in the vector 'x'
  x <- as.numeric(x)
  invalid_p_value <- which(as.numeric(x) < min_val | as.numeric(x) > max_val) # If the p-value is less than the minimum value, or more than the maximum value, the result was stored in invalid_p_value
  return(invalid_p_value) # Return invalid elements 
}
  
# Apply the function to p_values
invalid_p_value <- validate_p_value(normalised_gene_symbol$pvalue, 0, 1) # In the final combined data, looking at the p_value column, make sure the p-values are between 0 and 1

#Print the total number of invalid p-values 
cat("Total number of invalid p-values:", length(invalid_p_value))

# The total entries of invalid p-values is 813. This means that 813 rows are greater than 1 or lower than 0

# As there are 813 invalid rows, these were removed from the final combined data
removed_p_values <- normalised_gene_symbol[-invalid_p_value, ]

# Prints the final number of rows remaining after removing invalid cases
cat("Number of rows after removing invalid mouse strains:", nrow(removed_p_values))
# The output is 142066 rows

# Save the output file 
write.csv(removed_p_values, "~/Desktop/DCDM_GRP7/outputs/removed_p_values.csv", row.names = FALSE)
# The output file was saved, to be used for the next stage of cleaning 
```

### Validate the mouse strain to the 4 categories outlined in the SOP file 

```{r}
# Create a loop 
validate_mouse_strain <- function (x, field_name) { # Validate column values in the vector 'x'
  if (field_name == "mouse_strain") {
    valid_mouse_strains <- c("C57BL", "B6J", "C3H", "129SV") # Define a vector containing all valid mouse strains 
    invalid_mouse_strain <- which(!(x %in% valid_mouse_strains)) # Invalid mouse strains that do not match the vectors are stored in the invalid_mouse_strains
    return(invalid_mouse_strain) # Returns the invalid elements 
  }
}

# Apply the function to mouse strains 
invalid_mouse_strain <- validate_mouse_strain(removed_p_values$mouse_strain, "mouse_strain")

#Print the invalid rows 
print(invalid_mouse_strain)

# Print the total number of invalid mouse strains
cat("Total number of invalid mouse strain:", length(invalid_mouse_strain))

# The results show there are 1135 invalid mouse strain entries. There are 1135 strains that do not match the required stages in the SOP file

#Final combined data with the removed invalid entries 
clean_final_data <- removed_p_values[-invalid_mouse_strain, ]

#Prints the final number of rows remaining after removing invalid cases.
cat("Number of rows after removing invalid mouse strains:", nrow(clean_final_data))
# The output 140931 rows

# Save the output file, for further data cleaning
write.csv(clean_final_data, "~/Desktop/DCDM_GRP7/outputs/clean_final_data.csv", row.names = FALSE)
```

### Looking for Duplicates

```{r}
# Find duplicate rows
duplicates <- duplicated(clean_final_data)

# View duplicate rows
duplicate_rows <- clean_final_data[duplicates, ]
print(duplicate_rows)
# There are no duplicate rows

# Looking for duplicates in the unique identifier : analysis_id
duplicates_in_column <- clean_final_data[duplicated(clean_final_data$analysis_id), ]
print(duplicates_in_column)
```


# 2B: Cleaning the Disease information file

```{r}
# Install and load packages 
# install.packages("tibble")
library(tibble)
```

## Remove the header in bash
```{bash}
# Remove the first row (header) containing 'x'
sed '1D' Disease_information.txt > Disease_information_no_header.txt 

# Remove the numbers in quotation marks (e.g "1")
sed 's/"[0-9]*" "//g' Disease_information_no_header.txt > Disease_information_numbers_removed.txt 

# Remove unnecessary and inconsistent quotations across the file
sed 's/"//g' Disease_information_numbers_removed.txt > Disease_information_clean.txt 

# Using the SFTP command, the file was transferred out of the HPC onto our local project area
```

## Looping to assign correct columns
There are inconsistent commas in the CSV file (e.g some rows had 3 or 4 or 5 etc)
```{r}
# Loop to process each line
process_line <- function(line) {
  # Split the line by commas
  split_line <- strsplit(line, ",")[[1]]
  
  # Check if the line has enough parts (at least 3 columns)
  if (length(split_line) >= 3) {
    # First column: everything before the first comma
    first_column <- split_line[1]
    
    # Last column: everything after the last comma
    last_column <- split_line[length(split_line)]
    
    # Second-to-last column: everything between the second-to-last and last comma
    second_last_column <- split_line[length(split_line)-1]
    
    # Remainder as the second column: everything else in between
    second_column <- paste(split_line[2:(length(split_line)-2)], collapse = ",")
    
    # Return as a named vector
    return(c(first_column = first_column, second_column = second_column, 
             second_last_column = second_last_column, last_column = last_column))
  } else {
    # If the line doesn't have enough columns, return as is
    return(c(first_column = split_line[1], second_column = "", 
             second_last_column = "", last_column = ""))
  }
}

# Read the file line by line
lines <- readLines("~/Desktop/DCDM_GRP7/outputs/Disease_information_clean.txt")

# Apply the function to each line and store results in a data frame
clean_disease <- t(sapply(lines, process_line))

# Convert the result to a data frame
clean_disease <- as.data.frame(clean_disease)
```

## Convert to tibble for easy visualisation

```{r}
clean_disease_correct_row_names = as_tibble(clean_disease, rownames = "disease_id")

#remove first column
clean_disease_correct_row_names2 = clean_disease_correct_row_names[, -1]
```

## Rename the header with the correct column ids

```{r}
# Set the first row as the column names
colnames(clean_disease_correct_row_names2) = clean_disease[1, ]

# Remove the first row which is now the header
clean_disease_info = clean_disease_correct_row_names2[-1, ]  

# Check for missing values
sum(is.na(clean_disease_info)) 

#View cleaned file
View(clean_disease_info)

# Save the output as a CSV file 
write.csv(clean_disease_info, "~/Desktop/DCDM_GRP7/outputs/Disease_information_clean.csv", row.names = FALSE)

```

# 2C: Cleaning Parameter Description File

```{bash}
# Remove the first row (header) containing 'x'
sed '1D' IMPC_parameter_description.txt > IMPC_parameter_description_no_header.txt

# Remove the numbers in quotation marks (e.g "1")
sed 's/"[0-9]*" "//g' IMPC_parameter_description_no_header.txt > IMPC_parameter_description_numbers_removed.txt

# Rename column header 
sed -i '1s/.*/impcParameterOrigId, name, description, parameter_id/' IMPC_parameter_description_numbers_removed.txt  

# Remove unnecessary and inconsistent quotations across the file
sed 's/"//g' IMPC_parameter_description_numbers_removed.txt > IMPC_parameter_description_clean.txt

# Using the SFTP command, the file was transferred out of the HPC onto our local project area
```

##Looping to assign correct columns

```{r}
# Loop to process each line
process_line2 <- function(line) {
  # Split the line by commas
  split_line <- strsplit(line, ",")[[1]]
  
  # Check if the line has enough parts (at least 3 columns)
  if (length(split_line) >= 3) {
    # First column: everything before the first comma
    first_column <- split_line[1]
    
    # Last column: everything after the last comma
    last_column <- split_line[length(split_line)]
    
    # Second-to-last column: everything between the second-to-last and last comma
    second_last_column <- split_line[length(split_line)-1]
    
    # Remainder as the second column: everything else in between
    second_column <- paste(split_line[2:(length(split_line)-2)], collapse = ",")
    
    # Return as a named vector
    return(c(first_column = first_column, second_column = second_column, 
             second_last_column = second_last_column, last_column = last_column))
  } else {
    # If the line doesn't have enough columns, return as is
    return(c(first_column = split_line[1], second_column = "", 
             second_last_column = "", last_column = ""))
  }
}

# Read the file line by line
lines2 <- readLines("~/Desktop/DCDM_GRP7/outputs/IMPC_parameter_description_clean.txt")

# Apply the function to each line and store results in a data frame
clean_PD <- t(sapply(lines2, process_line2))

# Convert the result to a data frame
clean_PD <- as.data.frame(clean_PD)
```

## Convert to tibble for easy visualisation

```{r}
clean_PD_correct_row_names = as_tibble(clean_PD, rownames = "impcParameterOrigId")

# Remove first column
clean_PD_correct_row_names = clean_PD_correct_row_names[, -1]
```

## Rename the header with the correct column ids

```{r}
 # Set the first row as the column names
colnames(clean_PD_correct_row_names) = clean_PD_correct_row_names[1, ]  

# Remove the first row which is now the header
clean_parameter_description = clean_PD_correct_row_names[-1, ]            

# When we viewed the dataset, there were empty cells in the description column

# Replace missing values with NA

# Sets NA values to empty cells 
clean_parameter_description[clean_parameter_description == " " ] <- NA 

# Sets NA values into a string
clean_parameter_description[is.na(clean_parameter_description)] <- "NA" 

#Remove rows with inconsistent data
pd <- clean_parameter_description
filtered_data <- pd[grepl("^[0-9]+", pd[[1]]), ]

# Combine the header with the filtered data
clean_parameter_description <- rbind(pd[1, ], filtered_data)

#View cleaned data 
View(clean_parameter_description)

# Check for missing values
sum(is.na(clean_parameter_description)) 

# Save the output file 
write.csv(clean_parameter_description, "~/Desktop/DCDM_GRP7/outputs/IMPC_parameter_description_clean.csv", row.names = FALSE)
```

# 2D: Cleaning Procedure file

```{bash}
# Remove the numbers in quotation marks (e.g "1")
sed 's/"[0-9]*" "//g' IMPC_procedure.txt > IMPC_procedure_numbers_removed.txt

# Remove the column header, and remove the procedureID column as it contained no values
sed -i '1s/.*/name, description, isMandatory, impcParameterOrigId/' IMPC_procedure_numbers_removed.txt 

# Remove unnecessary and inconsistent quotations across the file
sed 's/"//g' IMPC_procedure_numbers_removed.txt > IMPC_procedure_clean.txt  

# Using the SFTP command, the file was transferred out of the HPC onto our local project area
```

## Looping to assign correct columns

```{r}
# Loop to process each line
process_line3 <- function(line) {
  # Split the line by commas
  split_line <- strsplit(line, ",")[[1]]
  
  # Check if the line has enough parts (at least 3 columns)
  if (length(split_line) >= 3) {
    # First column: everything before the first comma
    first_column <- split_line[1]
    
    # Last column: everything after the last comma
    last_column <- split_line[length(split_line)]
    
    # Second-to-last column: everything between the second-to-last and last comma
    second_last_column <- split_line[length(split_line)-1]
    
    # Remainder as the second column: everything else in between
    second_column <- paste(split_line[2:(length(split_line)-2)], collapse = ",")
    
    # Return as a named vector
    return(c(first_column = first_column, second_column = second_column, 
             second_last_column = second_last_column, last_column = last_column))
  } else {
    # If the line doesn't have enough columns, return as is
    return(c(first_column = split_line[1], second_column = "", 
             second_last_column = "", last_column = ""))
  }
}

# Read the file line by line
lines3 <- readLines("~/Desktop/DCDM_GRP7/outputs/IMPC_procedure_clean.txt")

# Apply the function to each line and store results in a data frame
clean_Pro <- t(sapply(lines3, process_line3))

# Convert the result to a data frame
clean_Pro <- as.data.frame(clean_Pro)
```

## Convert to tibble for easy visualisation

```{r}
# Changes rowname to correct first column name
clean_Pro_correct_row_names = as_tibble(clean_Pro, rownames = "name") 

# Remove the first column
clean_Pro_correct_row_names = clean_Pro_correct_row_names[, -1]
```

## Rename the header with the correct column ids

```{r}
 # Set the first row as the column names
colnames(clean_Pro_correct_row_names) = clean_Pro_correct_row_names[1, ]  

# Remove the first row which is now the header
clean_Pro = clean_Pro_correct_row_names[-1, ]    

# View the updated dataframe
View(clean_Pro)
# When we viewed the dataset, there were empty cells in the description column

# Replace NA values in a specific column with the string "NA"
clean_Pro[clean_Pro == ' '] <- NA
clean_Pro[is.na(clean_Pro)] <- "NA"

sum(is.na(clean_Pro)) #check for missing values

#View cleaned data
View(clean_Pro)
# Save the output file 
write.csv(clean_Pro, "~/Desktop/DCDM_GRP7/outputs/IMPC_procedure_clean.csv", row.names = FALSE)
```